---
title: "RU_tidyverse"
author: "Rockefeller University, Bioinformatics Resource Centre"
date: "https://rockefelleruniversity.github.io/RU_tidyverse_core/"
output: 
  html_document:
    number_sections: false  ## if you want number sections at each table header
    theme: united  # many options for theme, this one is my favorite.
    highlight: tango  # specifies the syntax highlighting style
params:
  toMessage: false    
---


<p>&nbsp;</p>
These exercises cover the sections of [tidyverse session](https://rockefelleruniversity.github.io/RU_tidyverse/). 

```{r, setup,include=FALSE}
toMessage <- ifelse(!is.null(params$toMessage),params$toMessage,FALSE)
message(toMessage)
```

1. Load in the tidyverse. Read in the fish ecology data in the csv file *exercises_fish.csv*, within the **data** directory. Read the file in again a second time but update the data types to reflect what they are i.e. 2 factors, an integer and a double. Save it as a variable called *fish_df*.

```{r,echo=toMessage}
library(tidyverse)

read_csv("data/exercises_fish.csv")

fish_df <- read_csv("data/exercises_fish.csv", col_types = cols(
  salmon_id = col_integer(),
  `common_name-size` = col_factor(),
  variable = col_factor(),
  value = col_double()))

fish_df
```

2. Tidy fish_df using pipes to link everything together. Update fish_df with this tidied dataframe. 

```{r,echo=toMessage}

fish_df %<>% 
  separate(`common_name-size`, sep = " - ", into=c("common_name", "age"), remove=TRUE) %>%
  pivot_wider(names_from = variable, values_from = value)

```

3. From your tidied fish_df create a dataframe with the variables _age_ and _IGF_ for only the _Steelhead_ fish

```{r,echo=toMessage}

fish_df %>% filter(common_name == 'Steelhead') %>%
  select( age, IGF1_ng_ml)
```

4. From your tidied fish_df create a dataframe with all the variables except the _IGF_ values, for all fish that begin with _S_. Arrange the dataframe based on common_name and length (largest first).

```{r,echo=toMessage}

fish_df %>% filter(str_starts(common_name,'S')) %>% 
  select( -IGF1_ng_ml) %>% arrange(common_name, desc(length_mm))
```

6. From your tidied fish_df perform these steps using a pipe:

- Group by the variable _age_. 
- Filter to get the biggest 5 by the variable _length_ in each group. 
- Summarize this data frame over the variable _length_ by calculating the mean. 

```{r}
fish_df %>% 
  group_by( age) %>%
  filter(rank(-length_mm) <= 5) %>% 
  summarise( mean_length_mm = mean(length_mm, na.rm = T))
```

7. From your tidied fish_df perform these steps using a pipe:
  - Filter to only look at _yearling_ in variable age
  - Group by common_name
  - Create new variable that is z-score of _length_ across groups (HINT: The *scale()* function can be used to calculate z-scores)
  - Create boxplot of grouped length z-scores
  
```{r,echo=toMessage}

fish_df %>% 
  filter(age == 'yearling') %>%
  group_by(common_name) %>% 
  mutate(length_zscore = scale(length_mm)) %>%
  ggplot(aes(x = common_name, y = length_zscore)) +
  geom_boxplot(fill=c('gold','darkorange'))
```


8. Read in the count data *tidy_counts.csv* from the **data** directory. Try setting different data types i.e. factors and logicals. What happens when we extract the vector of these counts? Data types are coerced if the wrong data types are specified.

```{r,echo=toMessage}

tidy_counts <- read_csv("data/tidy_counts.csv", col_types = cols(
    ENTREZ = col_character(),
    count_total = col_character(),
    Rep = col_logical()))

tidy_counts$ENTREZ %>% head(n=8)
tidy_counts$count_total %>% head(n=8)
tidy_counts$Rep %>% head(n=8)

```


9. Re-read the tidy counts CSV into R. How many ATPase genes (starting with ATP) are there and what are they? Subset dataframe to just these ATPase genes. Create a new lowercase variable that has 'ATP' removed from the symbol.


## Load in dataset and packages
```{r,echo=toMessage}
tidy_counts <- read_csv("data/tidy_counts.csv") 

tidy_counts %>% 
  pull(SYMBOL) %>%
  unique() %>% 
  str_starts('ATP') %>% 
  length()

tidy_counts %>% 
  pull(SYMBOL) %>%  
  str_starts('ATP') %>% 
  tidy_counts[.,] %>% 
  pull(SYMBOL) %>% 
  unique() 

tidy_counts %>%
  pull(SYMBOL) %>%  
  str_starts('ATP') %>% 
  tidy_counts[.,] %>%
  mutate(ATPtype = str_replace_all(SYMBOL,'ATP','' )) %>%
  mutate(ATPtype = str_to_lower(ATPtype))


```
